Supplementary Information

### 2020-11-28 doc/manuscript/MEE_manuscript.Rmd
knit with latest results
worked on manuscript

# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
#Appendices
site_details_appendix <- "S1"
DBH_reconstruction <- "S2"
climate_correction <- "S3"
traditional_comparison <- "S4"
changing_climate <- "S5"

site_details_appendix_title<- "Appendix S1. Methods for reconstruction of $DBH$"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4.  Methods for comparing climwin results with traditional methods"
changing_climate_title <- "Appendix S5.  Dealing with rapidly changing climate and tree growth"

@@

**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.** 
For dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.

@@

To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.  **HAVE YET TO TRY THIS **
For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
For models including $year$, we tested for first-order linear effects of $year$.
For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.

@@

Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`).
In the majority of cases (#%), the trend was negative.
Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).

@@

## changing growth rates
- growth rates changing directionally

@@

~~~~

@@

**The diversity of growth trends in relation to $DBH$ observed here (Fig. `r size`) is largely attributable to species ecology and stand history (Fig. 4).**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite Sheil et al. 2017 [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history. I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*

@@

`r interactions_and_year_summaries_legend`

>>>>

# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)

#Appendices
site_details_appendix <- "S1"
DBH_reconstruction <- "S2"
climate_correction <- "S3"
traditional_comparison <- "S4"
changing_climate <- "S5"
decadal_analysis_appendix <- "S6"

site_details_appendix_title<- "Appendix S1. Site Details"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4. Methods for comparing our approach with traditional methods"
changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"
decadal_analysis_appendix_title <- "Appendix S6. Analysis of DBH-growth relationships by decade"

@@

**Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; REF] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].**
However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.

@@

To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.  **(HAVE YET TO TRY THIS) **
For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
For models including $year$, we tested for first-order linear effects of $year$.
To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`). **(CAMERON'S WORK [IN THIS REPO](https://github.com/EcoClimLab/tree-rings_global-change))**
For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.

@@

There was a significant effect of `year` in the GLS models for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`), and these trends matched those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
In the majority of cases (#%), the growth trend over time was negative.
Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
Significant positive growth trends were observed for only # species, **LIST**, and all were modest compared to the steep negative trends observed for some species.
Growth rate was independent of `year` for only # species:  **LIST**.

@@

## changing growth rates
- growth rates changing directionally
- stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.

@@

- Analyses of CO~2~ effects reach inconsistent conclusions because of stand dynamics, sampling bias, analysis methods...
- We believe our analysis methods to be robust,
- Our sampling varies by sites, and sites analyzed have decent year x DBH coverage, but biases could remain (need to do some research on this, believe it's mostly stand dynamics)
- main bias would be stand dynamics. This is probably the case at secondary forests (HF, SCBI...). Even within mature forests, light-demanding species that establish in gaps (JACO at BCI, MEAZ at HKK) would tend to experience an increasingly competitive environment.

- In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
- At NM and SC, rapid declines in growth make sense in light of rapid climate change. They also correspond to observations of high tree mortality rates.
- At ZOF and CB,  trends are more mixed...
~~~~

@@

**The diversity of growth trends in relation to $DBH$ observed here (Fig. `r size`) is largely attributable to species ecology and stand history (Fig. 4).**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al. 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite Sheil et al. 2017 [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history. I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*

@@

### 2020-11-28 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

#Appendices
site_details_appendix <- "S1"
DBH_reconstruction <- "S2"
climate_correction <- "S3"
traditional_comparison <- "S4"
changing_climate <- "S5"

site_details_appendix_title<- "Appendix S1. Site Details"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4.  Methods for comparing our approach with traditional methods"
changing_climate_title <- "Appendix S5.  Dealing with rapidly changing climate and tree growth"

@@

>>>>

#Appendices
site_details_appendix <- "S1"
DBH_reconstruction <- "S2"
climate_correction <- "S3"
traditional_comparison <- "S4"
changing_climate <- "S5"
decadal_analysis_appendix <- "S6"

site_details_appendix_title<- "Appendix S1. Site Details"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4. Methods for comparing our approach with traditional methods"
changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"
decadal_analysis_appendix_title <- "Appendix S6. Analysis of DBH-growth relationships by decade"

@@

### Huai Kha Khaeng, Thailand

### Smithsonian Conservation Biology Institute, Virginia, USA

### Lilly Dickey Woods, Indiana, USA

### Harvard Forest, USA

### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic

### Niobrara, Nebraska, USA
Riparian site
BEPA are growing on N facing slopes near stream

### Little Tesuque, New Mexico, USA

### Cedar Breaks, Utah, USA

### Scotty Creek, Northwest Territories, Canada
melting permafrost
